---
title: "Getting Started"
format: html
---

## `{mlr3}` Basics

- Understand how `{mlr3}` is structured
- Access learners and (built-in) tasks
- Perform simple train-predict-evaluate steps


To get started, we load `{mlr3verse}`, which will load various packages from the `{mlr3}` ecosystem:

```{r}
library(mlr3verse)
```

`{mlr3}` ships with wrappers for many commonly used machine learning algorithms ("learners").  
We can access the list of available learners using the `mlr_learners` dictionary:

```{r}
as.data.table(mlr_learners)

lrn("classif.ranger")
```

There are also some built-in tasks, which we can access using the `mlr_tasks` dictionary:

```{r}
as.data.table(mlr_tasks)

tsk("penguins_simple")
```

Tasks encapsulate a data source (typically a `data.table`) and additional information regarding which variables are considered features and target. Tasks can also specify additional properties such as stratification, which we will see later.

### Example: Train-Predict-Evaluate

Let's perform a simple train-predict-evaluate step using a built-in task and learner for a simple classification example:

```{r}
task = tsk("penguins_simple")
learner = lrn("classif.ranger")

splits = partition(task, ratio = 0.7)

learner$train(task, row_ids = splits$train)
preds = learner$predict(task, row_ids = splits$test)
preds$score(msr("classif.ce"))
```

This code snippet trains a random forest model on the `penguins_simple` task (a simplified version of the `palmerpenguins` dataset, but without missing values) and evaluates the model's performance using the classification error metric.
